---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 2 - Due date 01/26/22"
author: "Vicky Janita"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is change "Student Name" on line 4 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp22.Rmd"). Submit this pdf using Sakai. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(readxl)
library(lubridate)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. The spreadsheet is ready to be used. Use the command $read.table()$ to import the data in R or $panda.read\_excel()$ in Python (note that you will need to import pandas package). }

```{r}
#Importing data set
mainData <- read_excel(path = "~/Library/Mobile Documents/com~apple~CloudDocs/VJ_Duke Documents/Spring 2022/ENV790.3 - Time Series/ENV790_TimeSeriesAnalysis_Sp2022/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip = 10 )


```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r, }

energyData <- mainData %>%
  select(c(1,4,5,6))
energyData <- energyData[-c(1),]
#head(energyData)

colnames(energyData) <-c("Month","BiomassProduction","RE_Production","HydroConsumption")

energyData$RE_Production <- as.numeric(energyData$RE_Production)
energyData$BiomassProduction <- as.numeric(energyData$BiomassProduction)
energyData$HydroConsumption <-as.numeric(energyData$HydroConsumption)

```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}

energyData$Month <- ymd(energyData$Month)
#energyData

myDate <- energyData$Month

ts_energyData <- ts(energyData[,2:4], frequency = 12, start = c(1973,1), end = c(2021,9))
#ts_energyData

```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
meanBiomassProduction <- mean(ts_energyData[,1])
meanREProduction <- mean(ts_energyData[,2])
meanHydroConsumption <- mean(ts_energyData[,3])

sdBiomassProduction <- sd(ts_energyData[,1])
sdREProduction <- sd(ts_energyData[,2])
sdHydroConsumption <- sd(ts_energyData[,3])
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}

plot(ts_energyData[,"RE_Production"],type ="l",col="blue",
     ylab = "Trillion BTU", ylim = c(100, 1100), 
     main ="Monthly Production or Consumption of Various Energy")
lines(ts_energyData[,"BiomassProduction"],col="red")
lines(ts_energyData[,"HydroConsumption"],col="green")
legend("topleft", legend=c("Total Renewable Energy Production", 
                           "Total Biomass Production", "Hydropower Electric Consumption"),
       col=c("blue", "red","green"), lty=1)

```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_energyData)
```

> Answer:
Hydroelectric Power consumption has a weak, negative correlation with Biomass and Renewable energy production (~ -0.28 and -0.06, respectively) while Biomass and Renewable Energy production shows a strong positive correlation at ~0.92 

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
biomass_ACF <- Acf(ts_energyData[,1], lag.max = 40, plot = FALSE)
plot(biomass_ACF, main = "Biomass Production ACF")
RE_ACF = Acf(ts_energyData[,2], lag.max = 40, plot = FALSE)
plot(RE_ACF, main = "Renewable Energy Production ACF")
hydro_ACF = Acf(ts_energyData[,3], lag.max = 40, plot = FALSE)
plot(hydro_ACF, main = "Hydroelectric Consumption ACF")
```

> Answer: 
Biomass and Renewable energy production shows a similar pattern where the ACF value continues to slowly decay from lag 1 to lag 40, all showing significance. However, the Hydro Consumption's ACF value indicates a seasonal pattern where it decays and grow again after each 6 lags. 


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
biomass_PACF = Pacf(ts_energyData[,1], lag.max = 40, plot = FALSE)
plot(biomass_PACF, main = "Biomass Production PACF")
RE_PACF = Pacf(ts_energyData[,2], lag.max = 40, plot = FALSE)
plot(RE_PACF, main = "Renewable Energy Production PACF")
hydro_PACF = Pacf(ts_energyData[,3], lag.max = 40, plot = FALSE)
plot(hydro_PACF, main = "Hydroelectric Consumption PACF")
```

>Answer: 
The PACF values for the biomass production and renewable energy production no longer as significance as the ACF values. In both variables, the PACF value significantly drops at lag 2 and several negative and significant PACF values between lag 1 to 40. Furthermore, the PACF value of hydroelectric power consumption no longer exhibits seasonal pattern. 


